---
title: "Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1. Loading and preprocessing the data

```{r}
library(ggplot2)
stepsdata <- read.csv(file="activity.csv", header=TRUE)
summary(stepsdata)
```

# Q2. What is mean total number of steps taken per day?
### 1. Calculate the total steps taken per day
```{r}
totalsteps <- sum(stepsdata$steps, na.rm = TRUE)
totalsteps
totalSteps <- aggregate(steps ~ date, stepsdata, FUN=sum)
```


### 2. Make a histogram
```{r echo=FALSE}
hist(totalSteps$steps,
     main = "Total Steps per Day",
     xlab = "Number of Steps")
```


### 3. Calculate Mean & Median
```{r}
meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
meanSteps
medSteps <- median(totalSteps$steps, na.rm = TRUE)
medSteps
```

# Q3. What is the average daily activity pattern

### 1. Make a time-series plot of the 5-minute interval and the average number of steps taken, averaged acoss all days.

```{r echo=FALSE}
meanStepsByInt <- aggregate(steps ~ interval, stepsdata, mean)

ggplot(data = meanStepsByInt, aes(x = interval, y = steps)) +
  geom_line() +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```


### 2. Which 5-minute interval across all days contain the maximum number of steps
```{r}
maxInt <- meanStepsByInt[which.max(meanStepsByInt$steps),]
maxInt
```

# Q4. Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset
```{r}
missingVals <- is.na(stepsdata$steps)
summary(missingVals)
```


### 2. Devise a strategy for filling in all of the missing values.
```{r}
#We use the strategy to replace each NA value by the mean of the steps attribute.

```

### 3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
newdata <- transform(stepsdata,
                              steps = ifelse(is.na(stepsdata$steps),
                                             meanStepsByInt$steps[match(stepsdata$interval, 
                                                                        meanStepsByInt$interval)],
                                             stepsdata$steps))
summary(newdata)
```


### 4.Make a histogram of the total number of steps taken each day and report the mean and median.
```{r}
impStepsByInt <- aggregate(steps ~ date, newdata, FUN=sum)
```


```{r echo=FALSE}
hist(impStepsByInt$steps,
     main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")
```

```{r}
impMeanSteps <- mean(impStepsByInt$steps, na.rm = TRUE)
impMedSteps <- median(impStepsByInt$steps, na.rm = TRUE)
#New Mean
impMeanSteps
#New Median
impMedSteps

#Difference in Mean
diffMean = impMeanSteps - meanSteps
diffMean
#Difference in Median
diffMed = impMedSteps - medSteps
diffMed
#Total Difference
diffTotal = sum(impStepsByInt$steps) - sum(totalSteps$steps)
diffTotal
```

# Q5. Are there differences in activity patterns between weekdays and weekends?
### 1. Create a new factor variable in the dataset with two levels - "weekend" and "weekday"
```{r}
DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
    return ("weekeday")
  else if (day %in% c('Saturday', 'Sunday'))
    return ("weekend")
  else
    stop ("Invalid Date Format.")
}
newdata$date <- as.Date(newdata$date)
newdata$day <- sapply(newdata$date, FUN = DayType)
```

### 2. Make a panel plot containnig a time-series plot of the 5-minute interval and the average number of steps taken across all weekdays or weekends
```{r}
meanStepsByDay <- aggregate(steps ~ interval + day, newdata, mean)
```

```{r echo=FALSE}
ggplot(data = meanStepsByDay, aes(x = interval, y = steps)) + 
  geom_line() +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Output created using the following code

##### Since the document is R Markdown v2 document, use function rmarkdown::render
##### rmarkdown::render("PA1_template.Rmd")
##### rmarkdown::render("PA1_template.Rmd" , output_format="md_document")
